### 2022-08-12 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@fa572cda4b73fa82f7f24e762cbdf281e286e661 🚀

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
directly queried from the database (\(N = 235\)). Most ecological
networks archived on Mangal are multilayer networks, i.e.~networks that
describe different types of interactions. We kept all networks whose
interactions were mainly of predation and herbivory types, and removed
the largest network (\(S = 714\)) for computational efficiency reasons.
Then, to this set we added food webs from two different sources: the New
Zealand dataset (\(N = 21\); Pomeranz et al. 2018) and the Tuesday Lake
dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our complete
dataset thus contained a total of \(257\) food webs. Of the New Zealand
and Tuesday Lake datasets, \(19\) networks had data on species' relative
abundances that were used in the neutral model (Box 2). These \(19\)
food webs constitute our \emph{abundance dataset}, which is a subset of
our complete dataset.

@@

\hypertarget{structure-of-maxent-networks}{%
\subsection{Structure of MaxEnt
networks}\label{structure-of-maxent-networks}}

We compared different measures of the structure of empirical food webs
with the ones of null, neutral, and MaxEnt networks
tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
Nestedness indicates how much the diet of specialist species is a subset
of the one of generalists, and is strongly associated with network
modularity (Fortuna et al. 2010). We also measured the maximum trophic
level \emph{maxtl}, network diameter \emph{diam} (i.e.~the longest of
the shortest paths between all species pairs; Albert and Barabasi 2002),
the average maximum similarity between species pairs \emph{MxSim}, the
proportion of cannibal species \emph{Cannib} (i.e.~the proportion of
self loops), and the proportion of omnivorous species \emph{Omniv}
(i.e.~species whose prey are of different trophic levels). \emph{MxSim},
\emph{Cannib}, and \emph{Omniv} are more deeply defined in Williams and
Martinez (2000).

Overall, we found that models based on the joint degree sequence (type
II null and MaxEnt network models) reproduced the structure of empirical
networks much better than the ones based on connectance and the neutral
model of relative abundances. On average, the type II MaxEnt network
model was better at predicting nestedness (\(0.62 \pm 0.08\)) than its
corresponding null model (\(0.73 \pm 0.05\); empirical networks:
\(0.63 \pm 0.09\)), as well as the proportion of cannibal species. This
might in part be due because nestedness was calculated using the
spectral radius of the adjacency matrix, which leverages information on
the network itself just like the MaxEnt network models. However, the
type II null model was better at predicting network diameter and average
maximum similarity between species pairs. Predictions were similar
between both type II models for the maximum trophic level and the
proportion of omnivorous species.

\hypertarget{tbl:measures_all}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_all}Standardized mean differences of
predicted network measures with empirical data for all networks
(\(n = 257\)). Positive (negative) values indicate that the measure is
overestimated (underestimated) on average. Empirical networks are most
food webs archived on Mangal, as well as the New Zealand and Tuesday
Lake food webs (our complete dataset). Null 1: Type I null model based
on connectance. MaxEnt 1: Type I MaxEnt network model based on
connectance. Null 2: Type II null model based on the joint degree
sequence. MaxEnt 2: Type II MaxEnt network model based on the joint
degree sequence. \(\rho\): nestedness measured by the spectral radius of
the adjacency matrix. \emph{maxtl}: maximum trophic level. \emph{diam}:
network diameter. \emph{MxSim}: average maximum similarity between
species pairs. \emph{Cannib}: proportion of cannibal species (self
loops). \emph{Omniv}: proportion of omnivorous species. \emph{entropy}:
SVD entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
null 1 & -0.167 & 0.980 & 1.428 & -0.502 & 2.007 & 1.493 &
0.056\tabularnewline
MaxEnt 1 & -0.226 & 0.831 & 1.274 & -0.524 & 1.982 & 1.863 &
0.106\tabularnewline
null 2 & 0.160 & -0.125 & 0.016 & 0.007 & 1.078 & 0.559 &
-0.023\tabularnewline
MaxEnt 2 & -0.015 & 0.178 & 0.565 & -0.282 & 0.698 & 0.589 &
0.058\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{tbl:measures_abund}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_abund}Standardized mean differences of
predicted network measures with empirical data for networks having
abundance data (\(n = 19\)). Positive (negative) values indicate that
the measure is overestimated (underestimated) on average. Empirical
networks are all New Zealand and Tuesday Lake food webs with abundance
data (our abundance dataset). Null 1: Type I null model based on
connectance. MaxEnt 1: Type I MaxEnt network model based on connectance.
Null 2: Type II null model based on the joint degree sequence. MaxEnt 2:
Type II MaxEnt network model based on the joint degree sequence.
\(\rho\): nestedness measured by the spectral radius of the adjacency
matrix. \emph{maxtl}: maximum trophic level. \emph{diam}: network
diameter. \emph{MxSim}: average maximum similarity between species
pairs. \emph{Cannib}: proportion of cannibal species (self loops).
\emph{Omniv}: proportion of omnivorous species. \emph{entropy}: SVD
entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
neutral & 0.367 & -0.090 & 0.027 & 0.266 & 6.870 & 0.576 &
-0.083\tabularnewline
null 1 & -0.134 & 0.950 & 1.919 & -0.369 & 2.077 & 0.614 &
0.068\tabularnewline
MaxEnt 1 & -0.229 & 1.020 & 1.946 & -0.355 & 2.215 & 0.801 &
0.121\tabularnewline
null 2 & 0.128 & -0.115 & -0.135 & 0.157 & 1.444 & 0.029 &
-0.021\tabularnewline
MaxEnt 2 & -0.010 & 0.054 & 0.243 & -0.062 & -0.038 & 0.083 &
0.038\tabularnewline
\bottomrule
\end{longtable}

The picture slightly changes when we consider another important property
of ecological networks, i.e.~their motifs profile. We measured the
proportion of three-species motifs, which can be considered as simple
building blocks of ecological networks (Milo et al. 2002; Stouffer et
al. 2007). Motifs are the backbone of complex ecological networks from
which network structure is built upon and play a crucial role in
community dynamics and assembly (Bastolla et al. 2009; Stouffer and
Bascompte 2011). In fig.~\ref{fig:motifs}, we show that the motifs
profile of networks generated using the type II MaxEnt network model was
very close to the one of empirical data, and that they made better
predictions than the type II null model. This is also shown in
fig.~\ref{fig:motifs_rel}, where we see that relationships between
motifs pairs of empirical food webs are very similar to the ones of the
type II MaxEnt networks. Overall, these results suggest that the type II
MaxEnt network model can reproduce many aspects of network structure.
This highlights the importance of the joint degree sequence in shaping
ecological networks.

@@

Moreover, we found that empirical networks are close to their maximum
entropy considering a fixed joint degree sequence (fig.~S3). Empirical
networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
between the SVD entropy of empirical food webs and their maximum entropy
configuration is plotted in the last panel of fig.~\ref{fig:measures}.
We also found no relationship between the difference in SVD entropy
between empirical and type II MaxEnt networks and species richness, the
number of interactions, or connectance (fig.~S4). Similarly, we found no
correlation between the difference in SVD entropy and the Jaccard
distance of empirical and type II MaxEnt networks.

Finally, in fig.~\ref{fig:measures} we also show how well empirical
measures are predicted using the type II MaxEnt network model. In
accordance with our previous results, we found that nestedness was very
well predicted by our model. However, the type II MaxEnt network model
overestimated the maximum trophic level and network diameter, especially
for networks with high empirical values. In fig.~S5, we show that the
pairwise relationships between these four measures and species richness
in empirical food webs are similar (in magnitude and sign) to the ones
in type II MaxEnt networks.

@@

@@

\leavevmode\hypertarget{ref-Bastolla2009ArcMut}{}%
Bastolla, Ugo, Miguel A. Fortuna, Alberto Pascual-Garcia, Antonio
Ferrera, Bartolo Luque, and Jordi Bascompte. 2009. {``The Architecture
of Mutualistic Networks Minimizes Competition and Increases
Biodiversity.''} \emph{Nature} 458 (7241): 1018--U91.
\url{https://doi.org/10.1038/nature07950}.

@@

@@

@@

>>>>

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
directly queried from the database (\(N = 235\)). Most ecological
networks archived on Mangal are multilayer networks, i.e.~networks that
describe different types of interactions. We kept all networks whose
interactions were mainly of predation and herbivory types, and removed
the largest network (\(S = 714\)) for computational efficiency reasons.
Then, to this set we added food webs from two different sources: the New
Zealand dataset (\(N = 21\); J. P. Pomeranz et al. 2018) and the Tuesday
Lake dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our
complete dataset thus contained a total of \(257\) food webs. Of the New
Zealand and Tuesday Lake datasets, \(19\) networks had data on species'
relative abundances that were used in the neutral model (Box 2). These
\(19\) food webs constitute our \emph{abundance dataset}, which is a
subset of our complete dataset.

@@

\hypertarget{structure-of-maxent-food-webs}{%
\subsection{Structure of MaxEnt food
webs}\label{structure-of-maxent-food-webs}}

We measured various properties of these configurations of maximum
entropy to evaluate how well they reproduce the structure of sampled
food webs. Specifically, we evaluated their nestedness \(\rho\), their
maximum trophic level \(maxtl\), their network diameter \(diam\), their
average maximum similarity between species pairs \(MxSim\) (Williams and
Martinez 2000), their proportion of cannibal species \(Cannib\), their
proportion of omnivorous species \(Omniv\), their SVD entropy, and their
motifs profile. Nestedness indicates how much the diet of specialist
species is a subset of the one of generalists (Delmas et al. 2019) and
was measured using the spectral radius of the adjacency matrix
(Staniczenko, Kopp, and Allesina 2013). In turn, the network diameter
represents the longest of the shortest paths between all species pair
(Albert and Barabasi 2002). Further, cannibal species are species that
can eat members of their own species (i.e.~species having self loops),
whereas omnivorous species can prey on different trophic levels
(Williams and Martinez 2000). Finally, a motifs profile represents the
proportion of three-species motifs (subnetworks), which can be
considered as simple building blocks of ecological networks (Milo et al.
2002; Stouffer et al. 2007). All of these properties are relatively easy
to measure and were chosen based on their ecological importance and
prevalent use in network ecology (McCann 2011; Delmas et al. 2019).

We compared the performance of both heuristic MaxEnt models in
predicting these measures to the one of the null
(tbl.~\ref{tbl:measures_all}) and neutral models
(tbl.~\ref{tbl:measures_abund}). Overall, we found that the models based
on the joint degree sequence (i.e.~the type II null and heuristic MaxEnt
models) reproduced the structure of empirical food webs much better than
the ones based on connectance (i.e.~the type I null and heuristic MaxEnt
models). This suggests that the predictive power of connectance might be
more limited than what was previously suggested (Timothée Poisot and
Gravel 2014). On the other hand, the neutral model of relative
abundances was surprisingly good at predicting the maximum trophic level
and the network diameter. However, with the exception of the network
diameter, the type II heuristic MaxEnt model was better at predicting
network structure than the neutral model for all measures considered.
This might be because, although neutral processes are important, they
act in concert with niche processes in determining species interactions
Canard et al. (2014). The joint degree sequence encodes information on
both neutral and niche processes because the number of prey and
predators a species has is determined by its relative abundance and
biological traits. Therefore, these results show that having information
on the number of prey and predators for each species substantially
improves the prediction of food-web structure, both compared to models
solely based on connectance and the ones solely based on species
relative abundances.

Next, the predictions of the type II heuristic MaxEnt model can be
compared to its null model counterpart. On average, the type II
heuristic MaxEnt model was better at predicting nestedness
(\(0.62 \pm 0.08\)) than its corresponding null model
(\(0.73 \pm 0.05\); empirical networks: \(0.63 \pm 0.09\)) for all
networks in our complete dataset (tbl.~\ref{tbl:measures_all}). This
might in part be due to the fact that nestedness was calculated using
the spectral radius of the adjacency matrix, which directly leverages
information on the network itself just like the heuristic MaxEnt models.
The proportion of self-loops (cannibal species) was also better
predicted by the type II heuristic MaxEnt model in comparison to the
type II null model. However, the type II null model was better at
predicting network diameter and average maximum similarity between
species pairs, and predictions of the maximum trophic level and the
proportion of omnivorous species were similar between both types of
models. This suggests that, although the joint degree sequence is an
important driver of food-web structure, other ecological constraints
might need to be added to our MaxEnt model to account for some emerging
food-web properties, especially the ones regarding food-chain lengths.

\hypertarget{tbl:measures_all}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_all}Standardized mean difference between
predicted network measures and empirical data for all food webs in our
complete dataset (\(N = 257\)). Positive (negative) values indicate that
the measure is overestimated (underestimated) on average. Empirical
networks include most food webs archived on Mangal, as well as the New
Zealand and Tuesday Lake food webs. Null 1: Type I null model based on
connectance. MaxEnt 1: Type I heuristic MaxEnt model based on
connectance. Null 2: Type II null model based on the joint degree
sequence. MaxEnt 2: Type II heuristic MaxEnt model based on the joint
degree sequence. \(\rho\): nestedness measured by the spectral radius of
the adjacency matrix. \(maxtl\): maximum trophic level. \(diam\):
network diameter. \(MxSim\): average maximum similarity between species
pairs. \(Cannib\): proportion of cannibal species (self loops).
\(Omniv\): proportion of omnivorous species. \emph{entropy}: SVD
entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
null 1 & -0.167 & 0.980 & 1.428 & -0.502 & 2.007 & 1.493 &
0.056\tabularnewline
MaxEnt 1 & -0.226 & 0.831 & 1.274 & -0.524 & 1.982 & 1.863 &
0.106\tabularnewline
null 2 & 0.160 & -0.125 & 0.016 & 0.007 & 1.078 & 0.559 &
-0.023\tabularnewline
MaxEnt 2 & -0.015 & 0.178 & 0.565 & -0.282 & 0.698 & 0.589 &
0.058\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{tbl:measures_abund}{}
\begin{longtable}[]{@{}rrrrrrrr@{}}
\caption{\label{tbl:measures_abund}Standardized mean difference between
predicted network measures and empirical data for all food webs in our
abundance dataset (\(N = 19\)). Positive (negative) values indicate that
the measure is overestimated (underestimated) on average. Empirical
networks include the New Zealand and Tuesday Lake food webs having
abundance data. Neutral: Neutral model of relative abundances. Null 1:
Type I null model based on connectance. MaxEnt 1: Type I heuristic
MaxEnt model based on connectance. Null 2: Type II null model based on
the joint degree sequence. MaxEnt 2: Type II heuristic MaxEnt model
based on the joint degree sequence. \(\rho\): nestedness measured by the
spectral radius of the adjacency matrix. \(maxtl\): maximum trophic
level. \(diam\): network diameter. \(MxSim\): average maximum similarity
between species pairs. \(Cannib\): proportion of cannibal species (self
loops). \(Omniv\): proportion of omnivorous species. \emph{entropy}: SVD
entropy.}\tabularnewline
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endfirsthead
\toprule
model & rho & maxtl & diam & MxSim & Cannib & Omniv &
entropy\tabularnewline
\midrule
\endhead
neutral & 0.367 & -0.090 & 0.027 & 0.266 & 6.870 & 0.576 &
-0.083\tabularnewline
null 1 & -0.134 & 0.950 & 1.919 & -0.369 & 2.077 & 0.614 &
0.068\tabularnewline
MaxEnt 1 & -0.229 & 1.020 & 1.946 & -0.355 & 2.215 & 0.801 &
0.121\tabularnewline
null 2 & 0.128 & -0.115 & -0.135 & 0.157 & 1.444 & 0.029 &
-0.021\tabularnewline
MaxEnt 2 & -0.010 & 0.054 & 0.243 & -0.062 & -0.038 & 0.083 &
0.038\tabularnewline
\bottomrule
\end{longtable}

Finally, in fig.~\ref{fig:measures} we also show how well empirical
measures are predicted using the type II MaxEnt network model. In
accordance with our previous results, we found that nestedness was very
well predicted by our model. However, the type II MaxEnt network model
overestimated the maximum trophic level and network diameter, especially
for networks with high empirical values. In fig.~S3, we show that the
pairwise relationships between these four measures and species richness
in empirical food webs are similar (in magnitude and sign) to the ones
in type II MaxEnt networks.

Moreover, we found that empirical networks are close to their maximum
entropy considering a fixed joint degree sequence (fig.~S4). Empirical
networks had an SVD entropy of \(0.89 \pm 0.04\), compared to type II
MaxEnt networks which had values of \(0.94 \pm 0.03\). The relationship
between the SVD entropy of empirical food webs and their maximum entropy
configuration is plotted in the last panel of fig.~\ref{fig:measures}.
We also found no relationship between the difference in SVD entropy
between empirical and type II MaxEnt networks and species richness, the
number of interactions, or connectance (fig.~S5). Similarly, we found no
correlation between the difference in SVD entropy and the Jaccard
distance of empirical and type II MaxEnt networks.

The picture slightly changes when we consider another important property
of ecological networks, i.e.~their motifs profile. We measured the
proportion of three-species motifs, which can be considered as simple
building blocks of ecological networks (Milo et al. 2002; Stouffer et
al. 2007). Motifs are the backbone of complex ecological networks from
which network structure is built upon and play a crucial role in
community dynamics and assembly (Stouffer and Bascompte 2011). In
fig.~\ref{fig:motifs}, we show that the motifs profile of networks
generated using the type II MaxEnt network model was very close to the
one of empirical data, and that they made better predictions than the
type II null model. This is also shown in fig.~\ref{fig:motifs_rel},
where we see that relationships between motifs pairs of empirical food
webs are very similar to the ones of the type II MaxEnt networks.
Overall, these results suggest that the type II MaxEnt network model can
reproduce many aspects of network structure. This highlights the
importance of the joint degree sequence in shaping ecological networks.

@@

@@

\leavevmode\hypertarget{ref-Bartomeus2016ComFra}{}%
Bartomeus, Ignasi, Dominique Gravel, Jason M. Tylianakis, Marcelo A.
Aizen, Ian A. Dickie, and Maud Bernard-Verdier. 2016. {``A Common
Framework for Identifying Linkage Rules Across Different Types of
Interactions.''} \emph{Functional Ecology} 30 (12): 1894--1903.
\url{https://doi.org/10.1111/1365-2435.12666}.

@@

@@

\leavevmode\hypertarget{ref-Canard2014EmpEva}{}%
Canard, E. F., N. Mouquet, D. Mouillot, M. Stanko, D. Miklisova, and D.
Gravel. 2014. {``Empirical Evaluation of Neutral Interactions in
Host-Parasite Networks.''} \emph{The American Naturalist} 183 (4):
468--79. \url{https://doi.org/10.1086/675363}.

@@

\leavevmode\hypertarget{ref-McCann2011FooWeb}{}%
McCann, Kevin S. 2011. \emph{Food Webs (MPB-50)}. \emph{Food Webs
(MPB-50)}. Princeton University Press.
\url{https://doi.org/10.1515/9781400840687}.

@@

\leavevmode\hypertarget{ref-Poisot2015SpeWhya}{}%
Poisot, Timothée, Daniel B. Stouffer, and Dominique Gravel. 2015.
{``Beyond Species: Why Ecological Interaction Networks Vary Through
Space and Time.''} \emph{Oikos} 124 (3): 243--51.
\url{https://doi.org/10.1111/oik.01719}.

\leavevmode\hypertarget{ref-Pomeranz2019InfPrea}{}%
Pomeranz, Justin P. F., Ross M. Thompson, Timothée Poisot, and Jon S.
Harding. 2019. {``Inferring Predatorprey Interactions in Food Webs.''}
\emph{Methods in Ecology and Evolution} 10 (3): 356--67.
\url{https://doi.org/10.1111/2041-210X.13125}.
